Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-11-15 index.qmd
section:issues; process:write

<!-- 1. What does CJ do? -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927; @Pollitt_2004; @Pollitt_2012a]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. Where are we now with DCJ? -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what are the problems of DCJ as a method? -->
Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed. This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role and influence of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

@@

<!-- 1. What is the general from of the CJ theory? -->
In its most general form, Thurstone's theory [-@Thurstone_1927] posits that when a judge compares two stimuli, the resulting dichotomous outcome depends on two main factors: the discriminal process and the law of comparative judgment. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus. @fig-discriminal_process shows a visual depiction of the discriminal process of two stimuli (objects).

![Discriminal processes of two stimuli (objects). Extracted from @Bramley_2008 [pp. 249] ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}

However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential. This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait. Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*. @fig-comparative_judgment shows a visual depiction of the discriminal difference of two stimuli (objects).

![Discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251] ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}

<!-- 2. But several simplifying assumption are made to ease its use -->
Thurstone’s general form of the theory primarily applies to pairwise comparisons of stimuli made by a single judge [@Thurstone_1927, pp. 267]. Therefore, to enhance the theory's practical applicability, he developed five distinct cases, each incorporating a series of increasingly simplifying assumptions. @fig-thurstone_cases outlines each case along with their key assumptions. For a detailed explanation of this theoretical progression, see @Thurstone_1927 and @Bramley_2008 [pp. 248-253].

<!-- 3. And yet Case V is is the most commonly used -->
Surprisingly, Case V is the most commonly used case in the CJ literature, despite incorporating the largest number of simplifying assumptions [@Bramley_2008, pp. 253]. This widespread adoption is primarily due to the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959], which offers a simpler statistical representation of the case. The BTL model closely aligns with Case V’s assumptions, with one key difference: the distribution of the discriminal process. While Case V assumes a Normal distribution, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). However, this adjustment does not significantly affect the model’s estimation or interpretation, as the logistic and normal distributions are related to each other by a factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16].

@@

<!-- 4. But two assumptions of Case V might not hold -->
However, because Case V was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927, pp. 269], prioritizing statistical ease of use over precision in trait measurement, two of its assumptions may be causing more problems than expected: the assumption of zero correlation between the stimuli, and the assumption of equal dispersion between stimuli.

<!-- @Thurstone_1927 justify the use of case V, on multiple assumptions, but the most important for our purpose are two: -->

<!--   (a) it assumes the correlation between stimuli is zero, this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects.  -->

<!--   Particularly on how CJ is used now, use the text example -->

@@

<!--   (b) it assumes the correlation between stimuli is zero, this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for.  -->

@@

<!-- (c) discriminal dispersions of the stimuli are equal, but it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

@@

<!-- "It is probably true that this variability of the discriminal dispersions on the psychological continuum is of relatively less serious importance in dealing with strictly homogenous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogenous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927]. -->

@@

## The disconnect between structural and measurement models

@@

<!-- 1. The BTL model is CJ's measurement model -->
As outlined in the previous section, the BTL model typically functions as the measurement model in a CJ study [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 2. But there is no structural model, only post estimation analyses -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests. For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what is the problem with this? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. 
<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->

<!-- 4. So, what is the solution?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for eah pair [@Thurstone_1927]. -->

>>>>

<!-- 1. Overview of CJ method in assessing traits -->
In *comparative judgment* (CJ) studies, judges assess a specific trait or attribute across various stimuli by performing pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. Each comparison produces a dichotomous outcome, indicating which stimulus is perceived to exhibit a higher trait level. For example, when assessing text quality, judges compare pairs of written texts (the stimuli) to determine the relative quality each text exhibit (the trait) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have emphasized three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability indicates that CJ requires a relatively small number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to produce trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. Furthermore, evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. On the other hand, research on validity suggests that scores generated by CJ can accurately represent the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Finally, research on practical applicability highlights the method's versatility across both educational and non-educational contexts [@Jones_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 3. But what critical issues remain CJ Research -->
Nevertheless, despite the increasing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unaddressed. This research primarily focuses on three: the over-reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data, the apparent disconnect between CJ's trait measurement and hypothesis testing, and the unclear role of comparison algorithms on the method's reliability and validity. The following sections will discuss each of these issues in detail, followed by the introduction of a theoretical model and its statistical translation, which aims to address all three concerns simultaneously.

@@

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory [-@Thurstone_1927b] posits that the dichotomous outcome resulting from comparing two stimuli is determined by two factors: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* refers to the psychological effect each stimulus has on the judges, or more simply stated, the judges’ perception of the trait level of each stimulus. Thurstone assumes that the discriminal process for each stimulus follows a Normal distribution. In this distribution, the mode (mean), known as the *modal discriminal process*, represents the position of the stimulus on the trait continuum, while the dispersion, known as the *discriminal dispersion*, reflects variability in the perceived trait level of the stimulus. @fig-discriminal_process shows example distributions of discriminal process for two stimuli (objects).

![Example distributions of discriminal processes for two stimuli (objects). Extracted from @Bramley_2008 [pp. 249]. ](/images/figures/discriminal_process.png){#fig-discriminal_process width=70%}

<!-- 2. Thurstone's Theory of CJ: discriminal difference -->
However, since the discriminal mode and dispersion of a single stimulus are not directly observable without comparing it to other stimuli, the *law of comparative judgment* becomes essential. This law asserts that when assessing a specific trait by comparing two stimuli, the stimulus positioned further along the continuum is perceived as having a higher level of that trait. Thus, the observed dichotomous outcome is determined by the distribution of the difference between the discriminal processes of the two stimuli, called the *discriminal difference*. @fig-comparative_judgment shows an example distribution of the discriminal difference for two stimuli (objects).

![Distribution of the discriminal difference between two stimuli (objects). Extracted from @Bramley_2008 [pp. 251]. ](/images/figures/comparative_judgment.png){#fig-comparative_judgment width=65%}

<!-- 3. Thurstone's Cases and their simplifying assumptions  -->
Importantly, the theory's general form primarily addresses pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. Consequently, to enhance its practical applicability, Thurstone introduced five distinct cases, each defined by progressively simplifying assumptions. @fig-thurstone_cases summarizes these cases, focusing on key assumptions such as the distribution of discriminal processes, the similarity of discriminal dispersion across stimuli, the correlation between stimuli, and which judges perform the comparisons. For a comprehensive discussion of this progression, refer to @Thurstone_1927b and @Bramley_2008 [pp. 248-253].

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Surprisingly, despite its reliance on the largest number of simplifying assumptions [@Bramley_2008, pp. 253], Case V remains the most widely used case in the CJ literature. This popularity is largely due to its simplified statistical representation in the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959]. The BTL model mirrors Case V’s assumptions, with one key difference: while Case V assumes a Normal distribution for the discriminal processes, the BTL model uses the more mathematically tractable logistic distribution [@Bramley_2008, pp. 254] (see @fig-thurstone_cases). This substitution has little effect on the model's estimation or interpretation, as the Normal and logistic distributions differ by a scaling factor of approximately $1.7$ [@vanderLinden_et_al_2017_I, pp. 16] (refer to @fig-logistic_vs_normal).

@@

::: {#fig-logistic_vs_normal layout-ncol=2}

![Probability density](/images/figures/density.png){#fig-density width=95%}

![Cummulative probability](/images/figures/cumulative.png){#fig-cumulative width=95%}

Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255].
:::

<!-- 2. But what are limitations of Case V (and the BTL model) -->
However, Case V (and its BTL model counterpart) was originally developed to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement. As a result, its assumptions may not be suitable for applications beyond the psychophysical contexts, which typically involve simple, homogeneous stimuli. Thurstone himself cautioned that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], further acknowledging that some assumptions could become problematic with less homogeneous stimuli, such as handwriting or English compositions [@Thurstone_1927a, pp. 374]. Consequently, given that current CJ applications often involve more complex, less homogeneous stimuli, two assumptions of Case V may not always hold theoretical or in practice: the assumption of zero correlation and equal dispersion between stimuli.

<!-- 3. The problems of assuming zero correlation between stimuli -->
[Regarding the assumption of *zero correlation between stimuli* ... ]{style="color:red;"}

<!-- this translates into the cancellation of judges effects by mean of opposing and equally weighted 'mood' and 'simultaneous contrasts' effects. This is demonstrated using the additive nature of the logit scale, which helps to cancel 'bias' judges effects. -->

@@

<!-- this also translates into the idea that stimuli are the main focus of estimation and analysis, but what happens when the focus of analysis is the individuals that generated those stimuli. Meaning there is an amount of correlation that it is not accounted for. example of Boonen -->

@@

<!-- "It is a safe assumption that when the stimulus series is very homogeneous with no distracting attributes, the correlation between discriminal deviations is low and possible even zero" [@Thurstone_1927b, pp. 268]. -->

<!-- 4. The problems of assuming equal dispersion between stimuli -->

<!-- it is not hard to imagine that certain individuals can produce good quality texts with more precision than other individuals. Example of exaggerated comparison of individuals, university students vs secondary school students. -->

@@

<!-- "It is probably true that this variability of the discriminal dispersion on the psychological continuum is of relatively less serious importance in dealing with strictly homogeneous stimulus series, but it becomes a serious factor in dealing with less conspicuous attributes or with less homogeneous stimulus series such as handwriting specimens, English compositions, sewing samples, oriental rugs." [@Thurstone_1927a, pp. 374]. -->

@@

## The disconnect between trait measurement and hypothesis testing

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model typically  functions as the measurement model for the trait under assessment [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of the texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

<!-- 2. Utilizing BTL Scores for further analysis and hypothesis testing -->
Researchers then typically use the estimated BTL scores, or their transformations, to conduct additional analyses or hypothesis tests. For example, these scores have been used to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
However, the statistical literature advises caution when using estimated scores to conduct additional analyses or hypotheses tests. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can introduce bias into the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of the bias are often unpredictable; results may be attenuated, exaggerated, or remain unaffected, depending on the amount of uncertainty present in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, reduced precision in hypothesis tests weakens their statistical power, ultimately increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. 
<!-- A type-I error occurs when a *true* null hypothesis is incorrectly rejected, while a type-II error occurs when a *false* null hypothesis is incorrectly accepted [@Everitt_et_al_2010]. -->

<!-- 4. So, how do we mitigate these risks?  -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

@@

<!-- It is assumed that the single observer compares each pair of stimuli a sufficient number of times so that a proportion may be determined for each pair [@Thurstone_1927b, pp. 267]. -->
